The Talking Heads Experiment

### 2015-04-16 chap8/chap8.tex
hyphenation, orphans, widows and some other layout updated

The ideas and results that have been discussed in Part I of the present volume 
were based on results obtained in idealised conditions. Computer simulations are able to 
test the validity of the basic algorithms, but simulations are still theoretical in nature. 
It is well known that technologies can only become a reality after 
experiments have been carried out in open-ended real world conditions. Nobody would want to fly an airplane that 
has only been tested in computer simulations.

We did the same for the Talking heads experiments and this was a huge step. Many more people had to get
get involved: for making the code more robust, setting up the teleportation infrastructure, and installing and 
maintaining the physical installations in different countries.  
The mechanisms about language evolution discussed in Part I of this book did not significantly change. On the contrary, 
they received solid experimental confirmation. At the same time, we learned a huge amount about the dependencies
between these mechanisms and the environments in which agents use them. We also learned a great deal about running software 
agents in an open infrastructure distributed over the entire globe. Furthermore we learned a lot 
about the interaction between humans and agents ranging from 
enthusiastic participation and enjoyment to nasty attacks by English hackers, set on destroying the experiment.

@@

Laboratorium will search the limits and possibilities of the places where knowledge and culture are made. Throughout 
the summer we will establish within the city of Antwerpen networks, fluctuating between highly specialized work by 
scientists, artists, dancers, and writers. "Working places" where the participants communicate their findings on the 
"work in progress". Also the scientific laboratories in Antwerpen will be involved in the initiative.

Laboratorium started as a discussion that involves questions such as: What is the meaning of laboratories? What is 
the meaning of experiments? When do experiments become public and when does the result of an experiment reach 
public consensus? Is rendering public what happens inside the laboratory of the scientist and the studio of the 
artist a contradiction in terms? These and other questions are being offered in this interdisciplinary project 
that starts with the "workplace" where the artists and the scientists experiment and work freely.}
\end{quotation}

@@

The exhibition took place in two locations: the photography museum in Antwerp, which was the main location, and an annex 
occupying several floors in a high rise office building close to the central station (the so called President's
Building). The Photography Museum Exhibition featured work by several well known artists, architects and scientists. 
The annex featured the Talking Heads experiment as well as 
work by artists Joseph Grigely and Matt Mullican and science philosopher 
Isabelle Stengers, who had set up a reconstruction of Galileo's famous experiments. 
There were also a number of public presentations under the heading 
"The Theatre of Proof", organised by Bruno Latour. The catalog was designed by Bruce Mau and his team.

@@

Months of preparation and testing took place before we attempted to go 'live' and public. 
Once this happened, work continued frantically to shake out bugs, get the semiotic dynamics right
and then maintain the general communication infrastructure and the ongoing interactions with participants. 
The fascinating email correspondence (see section 8.3) shows that most of the difficulties initially came 
from running the agent teleportation infrastructure. The reader has to keep in mind that this was 
late nineties, when large-scale uploading 
and downloading, cloud computing, agent architectures, etc. were totally in their infancy.

@@

The installation in Antwerp was anounced as a "Laboratory for Cognitive Robotics and Teleportation".\cite{Steels:99b}
It featured various rooms as shown in the layout in \figref{fig:layout}. The rooms had different functions: 
\begin{enumerate} 
\item The {\itshape central room} which was visible on entering the space contained the experiment itself, i.e. the 
two cameras mounted on tripods, the computers driving them, the screen with geometric figures, and a projection of 
what went on. The activities of the agents were audible through a narration. 
\item A {\itshape reading room} contained background philosophical and scientific 
papers and excerpts from books about language, meaning and their origins. These readings included: 
\item A {\itshape user interface room} contained a workstation where visitors to the exhibition could create their own 
agents, direct them to play games on certain sites, and teach new words to their agents. 
\end{enumerate}

\begin{figure}[htbp]
  \centerline{\includegraphics[width=0.9\textwidth]{chap8/figures/layout.pdf}}
\caption{\label{fig:layout}Layout of the 'Laboratory for Cognitive Robotics and Teleporation" at the Laboratorium exhibition in Antwerp.}
\end{figure}

\begin{figure}[htbp]
 \centerline{\includegraphics[width=.35\textwidth]{chap8/figures/zoom-in-head.pdf}}
\caption{\label{fig:laboratorium}A single Talking Head at the Paris site. In the background we see (bottom) the display with an
image that the agent picked up, (on top of that) a screen displaying the processed image and the discrimination trees, 
and to the right the loudspeaker broadcasting the speech output of the agent.}
\end{figure}

@@

Once the physical installation at the Antwerp site was completed, work focused on getting the experiment itself 
up and running. It is interesting and instructive to look at some snippets of the email correspondence during the 
first weeks. The despair when trying to cope with the unavoidable problems but also the excitement 
as a language began to emerge shines through. The English humor of Angus McIntyre was a welcome antidote to 
all the stress. The correspondence reprinted here is a small selection but also necessary fragmentary because a 
lot of communication took place face-to-face or through the telephone, as email was not always possible.

We pick up the email correspondence on
June 23, 1999 when the first experiments are taking place to try and make language games at the three sites possible 
and allow the scheduling and activation of agents.

@@

Enjoy 
A
\end{mail}
\begin{mail} 
--============\_-1278188520==\_d============\\
To: Joris Van Looveren <joris@arti.vub.ac.be>\\
From: Angus McIntyre <angus@csl.sony.fr>\\
Subject: Re: Brussels 2pm\\

@@

I'm off home now. If you need me to come back and kick 
the server, call me on:
	+33-1-42-78-73-88
Bye,
	A
\end{mail}
}

\subsection*{25 June 1999}

@@

Fred and Luc.
\end{mail}

\begin{mail}
--============\_-1278188520==\_d============\\
Date: Tue, 29 Jun 1999 23:25:45 +0200 (MET DST)\\
Subject: There is hope for robokind \\

@@

A total of 8000 words and 500 concepts were created by the agents, with a core vocabulary consisting of 100 basic words expressing concepts like up, down, left, right, green, red, large, small, etc. Of these, 8 words represent a large majority of words used (about 80 \%). 4 of these words refer to the position of objects: GOREWA (top), DOWN (bottom), WOGGLESPLAT (left), and SESUBIPU (right). 4 other words refer to colors: ROUGE (red), KAZOZO (green), WEGIRIRA (blue), and EMPTY (light). The distribution of these words after 130,000 games is shown in \figref{fig:words-used}.

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=\textwidth]{chap8/figures/wogglesplat.pdf}}
\caption{\label{fig:woggle}Meaning-Form diagram: Different words for expressing the meaning 'left', i.e. 
horizontal position is less than 0.25 (scaled). 
New words come up all the time but there is a clear winner-take-all effect with "wogglesplat" winning. 
}
\end{figure}

\figref{fig:droite} shows another example of the semiotic dynamics in the experiment, this time looking at all the 
meanings for a particular word, namely `droite' (meaning `to the right' in French). This word clearly has been introduced
by a human user and the dominating meaning progressively becomes a region (between 0.75 and 1.0) on the horizontal position, 
as could be expected. On the way we see some confusion, Specifically there must be objects that appear both on the right 
and up (vertical position between 0.5 and 1.0).

\begin{figure}[htbp]
  \centerline{\includegraphics[width=\textwidth]{chap8/figures/droite.pdf}}
\caption{\label{fig:droite}Form-Meaning diagram: Different meanings associated with the word `droite'. The dominant meaning is 
in line with the human use of the word, namely horizontal position (HPOS) to the right of the image.}
\end{figure}

\figref{fig:bozopite} shows different meanings competing for the same word "bozopite". Dominant ones are 
a large area (AREA scaled between 0.5 and 1.0) or extended width (WIDTH scaled between 0.5 and 1.0). These two concepts
apparently could both be applied in many situations, but at some point new situations appeared that 
caused a symmetry breaking and AREA became dominant.

@@

It was also possible that different meanings which were highly compatible were maintained in the population. 
An example is shown in the form-meaning diagram in \figref{fig:down}, which plots the average 
frequency of meanings for the word `down'. The meanings are all concepts on the vertical position channel 
VPOS, but they carve out smaller and smaller regions.

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.65\textwidth]{chap8/figures/down.pdf}}
\caption{\label{fig:down}Form-Meaning diagram: Different meanings associated with the word "down". There are in particular two meanings. 
one is a region on the red-channel and another one a region on the vertical position channel. 
}
\end{figure}

@@

The Talking Heads experiment was without doubt a success from many angles. The mechanisms for concept formation,
lexicon formation and alignment (as discussed in Part I of this book) all worked out the 
way it was expected, even in very difficult 'real world' conditions. The complex hardware could be maintained at the 
different physical sites and the general web infrastructure and agents teleportation mechanisms held up despite a 
significant scale-up. An enthusiastic user group formed and they actively created agents and sent 
them out on the network, often also teaching their agents human language words which then propagated
in the rest of the population. Users became very attached to their agents, upset when their agents could not get 
to the sites they had scheduled (because it took almost 1 minute for a complete language game and so other agents 
had to wait), and trying to figure out how and why they had learned certain words.

@@

The experiment ended with the following ``tongue-in-cheek'' post by Angus McIntyre on the Forum: \\

@@

Here is one commentary of a dedicated user: \\

>>>>

The ideas and results that have been discussed in \partref{part:1} of the present volume 
were based on results obtained in idealised conditions. Computer simulations are able to 
test the validity of the basic algorithms, but simulations are still theoretical in nature. 
It is well known that technologies can only become a reality after 
experiments have been carried out in open-ended real world conditions. Nobody would want to fly an airplane that 
has only been tested in computer simulations.

We did the same for the Talking heads experiments and this was a huge step. Many more people had to get
get involved: for making the code more robust, setting up the teleportation infrastructure, and installing and 
maintaining the physical installations in different countries.  
The mechanisms about language evolution discussed in \partref{part:1} of this book did not significantly change. On the contrary, 
they received solid experimental confirmation. At the same time, we learned a huge amount about the dependencies
between these mechanisms and the environments in which agents use them. We also learned a great deal about running software 
agents in an open infrastructure distributed over the entire globe. Furthermore we learned a lot 
about the interaction between humans and agents ranging from 
enthusiastic participation and enjoyment to nasty attacks by English hackers, set on destroying the experiment.

@@

Laboratorium will search the limits and possibilities of the places where knowledge and culture are made. Throughout 
the summer we will establish within the city of Antwerpen networks, fluctuating between highly specialized work by 
scientists, artists, dancers, and writers. ``Working places" where the participants communicate their findings on the 
``work in progress". Also the scientific laboratories in Antwerpen will be involved in the initiative.

Laboratorium started as a discussion that involves questions such as: What is the meaning of laboratories? What is 
the meaning of experiments? When do experiments become public and when does the result of an experiment reach 
public consensus? Is rendering public what happens inside the laboratory of the scientist and the studio of the 
artist a contradiction in terms? These and other questions are being offered in this interdisciplinary project 
that starts with the ``workplace" where the artists and the scientists experiment and work freely.}
\end{quotation}

@@

The exhibition took place in two locations: the photography museum in Antwerp, which was the main location, and an annex 
occupying several floors in a high rise office building close to the central station (the so called President's
Building). The Photography Museum Exhibition featured work by several well known artists, architects and scientists. 
The annex featured the Talking Heads experiment as well as 
work by artists Joseph Grigely and Matt Mullican and science philosopher 
Isabelle Stengers, who had set up a reconstruction of Galileo's famous experiments. 
There were also a number of public presentations under the heading 
``The Theatre of Proof", organised by Bruno Latour. The catalog was designed by Bruce Mau and his team.

@@

Months of preparation and testing took place before we attempted to go ``live'' and public. 
Once this happened, work continued frantically to shake out bugs, get the semiotic dynamics right
and then maintain the general communication infrastructure and the ongoing interactions with participants. 
The fascinating email correspondence (see section 8.3) shows that most of the difficulties initially came 
from running the agent teleportation infrastructure. The reader has to keep in mind that this was 
late nineties, when large-scale uploading 
and downloading, cloud computing, agent architectures, etc. were totally in their infancy.

@@

The installation in Antwerp was anounced as a ``Laboratory for Cognitive Robotics and Teleportation" \citep{Steels:99b}.
It featured various rooms as shown in the layout in \figref{fig:layout}. The rooms had different functions: 
\begin{enumerate} 
\item The {\itshape central room} which was visible on entering the space contained the experiment itself, i.e. the 
two cameras mounted on tripods, the computers driving them, the screen with geometric figures, and a projection of 
what went on. The activities of the agents were audible through a narration. 
\clearpage
\item A {\itshape reading room} contained background philosophical and scientific 
papers and excerpts from books about language, meaning and their origins. These readings included: 
\item A {\itshape user interface room} contained a workstation where visitors to the exhibition could create their own 
agents, direct them to play games on certain sites, and teach new words to their agents. 
\end{enumerate}

\begin{figure}[bp]
  \centerline{\frame{\includegraphics[width=0.9\textwidth]{chap8/figures/layout.pdf}}}
\caption{\label{fig:layout}Layout of the ``Laboratory for Cognitive Robotics and Teleporation" at the Laboratorium exhibition in Antwerp.}
\end{figure}

\todo{I have framed Figure 8.2 because it contains multiple images}
\begin{figure}[htbp]
 \centerline{\includegraphics[width=.35\textwidth]{chap8/figures/zoom-in-head.pdf}}
\caption{\label{fig:laboratorium}A single Talking Head at the Paris site. In the background we see (bottom) the display with an
image that the agent picked up, (on top of that) a screen displaying the processed image and the discrimination trees, 
and to the right the loudspeaker broadcasting the speech output of the agent.}
\end{figure}

@@

Once the physical installation at the Antwerp site was completed, work focused on getting the experiment itself 
up and running. It is interesting and instructive to look at some snippets of the email correspondence during the 
first weeks. The despair when trying to cope with the unavoidable problems but also the excitement 
as a language began to emerge shines through. The English humor of Angus McIntyre was a welcome antidote to 
all the stress. The correspondence reprinted here is a small selection but also necessary fragmentary because a 
lot of communication took place face-to-face or through the telephone, as email was not always possible. 
\enlargethispage{1\baselineskip}
We pick up the email correspondence on
June 23, 1999 when the first experiments are taking place to try and make language games at the three sites possible 
and allow the scheduling and activation of agents.

@@

Enjoy 
A
\end{mail}
\clearpage
\begin{mail} 
--============\_-1278188520==\_d============\\
To: Joris Van Looveren <joris@arti.vub.ac.be>\\
From: Angus McIntyre <angus@csl.sony.fr>\\
Subject: Re: Brussels 2pm\\

@@

I'm off home now. If you need me to come back and kick 
the server, call me on:
	+33-1-42-78-73-88
Bye,
	A
\end{mail}
}
\clearpage
\subsection*{25 June 1999}

@@

Fred and Luc.
\end{mail}
\clearpage
\begin{mail}
--============\_-1278188520==\_d============\\
Date: Tue, 29 Jun 1999 23:25:45 +0200 (MET DST)\\
Subject: There is hope for robokind \\

@@

A total of 8000 words and 500 concepts were created by the agents, with a core vocabulary consisting of 100 basic words expressing concepts like up, down, left, right, green, red, large, small, etc. Of these, 8 words represent a large majority of words used (about 80 \%). 4 of these words refer to the position of objects: GOREWA (top), \textsc{down} (bottom), WOGGLESPLAT (left), and SESUBIPU (right). 4 other words refer to colors: ROUGE (red), KAZOZO (green), WEGIRIRA (blue), and EMPTY (light). The distribution of these words after 130,000 games is shown in \figref{fig:words-used}.

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=\textwidth]{chap8/figures/wogglesplat.pdf}}
\caption{\label{fig:woggle}Meaning-Form diagram: Different words for expressing the meaning 'left', i.e. 
horizontal position is less than 0.25 (scaled). 
New words come up all the time but there is a clear winner-take-all effect with "wogglesplat" winning. 
}
\end{figure}
\enlargethispage{1\baselineskip}
\figref{fig:droite} shows another example of the semiotic dynamics in the experiment, this time looking at all the 
meanings for a particular word, namely `droite' (meaning `to the right' in French). This word clearly has been introduced
by a human user and the dominating meaning progressively becomes a region (between 0.75 and 1.0) on the horizontal position, 
as could be expected. On the way we see some confusion, Specifically there must be objects that appear both on the right 
and up (vertical position between 0.5 and 1.0).

\begin{figure}[htbp]
  \centerline{\includegraphics[width=\textwidth]{chap8/figures/droite.pdf}}
\caption{\label{fig:droite}Form-Meaning diagram: Different meanings associated with the word `droite'. The dominant meaning is 
in line with the human use of the word, namely horizontal position (\textsc{hpos}) to the right of the image.}
\end{figure}

\figref{fig:bozopite} shows different meanings competing for the same word "bozopite". Dominant ones are 
a large area (\textsc{area} scaled between 0.5 and 1.0) or extended width (\textsc{width} scaled between 0.5 and 1.0). These two concepts
apparently could both be applied in many situations, but at some point new situations appeared that 
caused a symmetry breaking and \textsc{area} became dominant.

@@

It was also possible that different meanings which were highly compatible were maintained in the population. 
An example is shown in the form-meaning diagram in \figref{fig:down}, which plots the average 
frequency of meanings for the word `down'. The meanings are all concepts on the vertical position channel 
\textsc{vpos}, but they carve out smaller and smaller regions.

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.65\textwidth]{chap8/figures/down.pdf}}
\caption{\label{fig:down}Form-Meaning diagram: Different meanings associated with the word ``down". There are in particular two meanings. 
one is a region on the red-channel and another one a region on the vertical position channel. 
}
\end{figure}

@@

The Talking Heads experiment was without doubt a success from many angles. The mechanisms for concept formation,
lexicon formation and alignment (as discussed in \partref{part:1} of this book) all worked out the 
way it was expected, even in very difficult 'real world' conditions. The complex hardware could be maintained at the 
different physical sites and the general web infrastructure and agents teleportation mechanisms held up despite a 
significant scale-up. An enthusiastic user group formed and they actively created agents and sent 
them out on the network, often also teaching their agents human language words which then propagated
in the rest of the population. Users became very attached to their agents, upset when their agents could not get 
to the sites they had scheduled (because it took almost 1 minute for a complete language game and so other agents 
had to wait), and trying to figure out how and why they had learned certain words.

@@

The experiment ended with the following ``tongue-in-cheek'' post by Angus McIntyre on the Forum:

@@

Here is one commentary of a dedicated user:
